Barriers to Green Living: Insights from 2018 and 2021 Toronto Climate Perception Surveys

### 2024-12-04 paper/climate_change.qmd
final touches on model

# Model {#sec-model}
TO DO:
The goal of our modelling strategy is twofold. Firstly,...
Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-appendix].
## Model set-up
In this analysis, we aim to predict the likelihood of taking climate-friendly actions based on demographic variables namely age and education. The specific actions include investing in an electric or hybrid vehicle instead of a gas-powered one, making home improvements (e.g., using LED lightbulbs), adopting meat alternatives, minimizing car use, purchasing green products, reducing hydro usage, reducing waste by repurposing used products, sorting waste correctly, and walking or cycling shorter distances.

### Model Specifications ###
TO DO:

@@

 - $y_i$ represents the likelihood of taking a climate-friendly action for individual $i$, measured on a five point Likert scale:
 1. "Very unlikely"
 2. "Somewhat unlikely"
 3. "Somewhat likely"
 4. "Very likely"
 5. "Already doing this or have done this"
- $\beta_0$ is the intercept, representing the baseline likelihood of taking action when all predictors are zero
- $\beta_1$ is the coefficient for Age, representing how a one-year increase affects the likelihood of taking action
- $\beta_2$ is the coefficient for Education, a categorical variable that is dummy-coded.
- $\epsilon_i$ is the error term, accounting for unexplained variation.

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model Justification 
This model is designed to explore how various demographics influence the likelihood of engaging in climate-friendly behaviors. The likelihood of taking specific actions, such as reducing energy consumption or investing in green technologies, represents an important outcome in understanding climate change perceptions. As such, focusing on these behaviors rather than general knowledge allows for a more actionable and directly applicable model to climate change policy and interventions.

#### Response Variable ####
The response variable in this analysis is the likelihood of taking climate-friendly actions, which is typically a continuous or ordinal measure based on respondents' self-reported likelihood of adopting behaviors such as switching to an electric vehicle, reducing waste, or conserving energy. These actions are commonly measured on a Likert scale (e.g., 1 = "not likely" to 5 = "very likely").

@@

Age: Previous studies indicate that younger individuals are more likely to engage in climate-positive behaviors due to greater environmental awareness and concern (Leiserowitz et al., 2010; Gifford, 2013).

Education: Higher educational attainment has been associated with a higher likelihood of adopting environmentally responsible behaviors, likely due to increased awareness of environmental issues and access to relevant information (Kollmuss & Agyeman, 2002).

Knowledge: A greater understanding of climate change is expected to positively influence individuals' likelihood of taking action. More knowledgeable individuals are more likely to adopt climate-friendly practices, as they are more aware of the urgency of the crisis and the potential benefits of sustainable behaviors (Gifford, 2013).

#### Model Structure ####
The model assumes that the likelihood of taking climate-friendly actions is linearly related to the input variables. Each predictor (age, education, and knowledge) is assumed to have a linear relationship with the likelihood of taking action, meaning that as these variables change, the likelihood of taking climate-friendly actions will either increase or decrease in a predictable manner. This structure facilitates a clear interpretation of how each variable contributes to the overall likelihood of climate-positive behavior.

Detailed Structure
The model does not assume any interactions between the predictors, meaning each factor (age, education, knowledge) is considered independently in its influence on the outcome. However, depending on the data, you might consider adding interaction terms if you believe that, for example, the effect of knowledge on action is stronger for younger individuals.
The linear relationship between predictors and the likelihood of action allows for easy interpretation, with coefficients representing the change in the likelihood of taking action for each unit change in the predictor variable.

#### Parameter Estimation ####
In parameter estimation, we anticipate that:
Age: We expect that younger individuals will have a higher likelihood of taking climate-friendly actions, as studies have shown that younger generations are more environmentally conscious and concerned about climate change (Leiserowitz et al., 2010).
Education: We hypothesize that higher levels of education will positively influence the likelihood of taking climate-positive actions. More educated individuals tend to have better access to information about climate change and sustainability practices (Kollmuss & Agyeman, 2002).
Knowledge: We expect that greater knowledge of climate change will be strongly associated with higher likelihoods of adopting climate-friendly behaviors. Knowledge acts as a critical enabler, motivating individuals to take action and contribute to mitigating climate change (Gifford, 2013).
We will estimate the model parameters using a standard linear regression approach (e.g., using the lm() function in R), which will provide the coefficients (
β
β) for each predictor, showing how each factor influences the likelihood of taking climate-friendly actions.

>>>>

# Model {#sec-model}
The goal of our modelling strategy is twofold. Firstly, we aim to predict the likelihood of individuals engaging in climate-friendly actions based on demographic variables, specifically age and education. The actions considered include investing in electric or hybrid vehicles, making home improvements (such as using LED lightbulbs), adopting meat alternatives, minimizing car use, purchasing green products, reducing hydro usage, reducing waste, sorting waste correctly, and walking or cycling shorter distances. Secondly, we aim to understand how these demographic characteristics influence individuals' likelihood of adopting these behaviors.

Here we describe the probability tree model used to investigate the relationship between likelihood and the demographic factors of age and education. Further details on background and diagnostics can be found in [Appendix -@sec-appendix].

## Model set-up
In this analysis, we focus on predicting the likelihood of taking climate-friendly actions based on the demographic variables of age and education. The likelihood is measured on a five-point Likert scale, ranging from "Very unlikely" to "Already doing this or have done this."

### Model Specifications ###
We use a probability tree model to predict the likelihood of individuals engaging in climate-friendly actions based on two key demographic variables: age and education.

The general form of the model is:

@@

 - $y_i$ represents the likelihood of individual $i$ engaging in a climate-friendly actin, measured on a five point Likert scale:
 1. "Very unlikely"
 2. "Somewhat unlikely"
 3. "Somewhat likely"
 4. "Very likely"
 5. "Already doing this or have done this"
- $\beta_0$ is the intercept, representing the baseline likelihood of taking action when all predictors are zero.
- $\beta_1$ is the coefficient for Age, representing how a one-year increase affects the likelihood of taking action.
- $\beta_2$ is the coefficient for Education, a categorical variable that is dummy-coded.
- $\epsilon_i$ is the error term, accounting for unexplained variation.

This model focuses on the direct effect of age and education on the likelihood of taking climate-friendly actions, assuming a linear relationship between these demographic variables and the outcome variable (the likelihood of action).

### Model Justification 
This model was developed to explore how demographic factors such as age and education influence individuals' likelihood of engaging in climate-friendly behaviors. By focusing on specific climate actions like reducing energy consumption, adopting sustainable practices, and investing in green technologies, the model aims to provide actionable insights for climate change interventions.

#### Response Variable ####
The response variable in this analysis is the likelihood of taking climate-friendly actions, measured using a Likert scale (1 = "Very unlikely" to 5 = "Already doing this or have done this"). This scale captures the respondent’s perceived likelihood of engaging in various climate-friendly behaviors

@@

- **Age**: Previous research indicates that younger individuals may be more inclined to adopt climate-friendly behaviors due to increased environmental awareness (Leiserowitz et al., 2010; Gifford, 2013).
- **Education**: Higher levels of education are often linked to greater awareness of environmental issues and a higher likelihood of adopting sustainable behaviors (Kollmuss & Agyeman, 2002). Education is treated as a categorical variable and dummy-coded for analysis.

#### Model Structure ####
This model assumes a linear relationship between the predictors (age and education) and the likelihood of taking action. Specifically, the likelihood of taking climate-friendly actions is modeled as a function of age and education, with coefficients representing the change in the likelihood of action associated with a unit change in each predictor.

There are no interaction terms in this model, meaning we assume that the effects of age and education on the likelihood of taking action are independent of each other. This simplicity ensures clear interpretation of each predictor's individual effect on the likelihood of taking climate-friendly actions.

## Nine Models ##

We fit nine separate probability tree models, one for each climate-friendly action we are predicting. Each model uses age and education as the predictors to estimate the likelihood of the corresponding action. These actions are:

1. Likelihood of investing in an electric or hybrid vehicle
2. Likelihood of making home improvements (e.g., using LED lightbulbs)
3. Likelihood of adopting meat alternatives
4. Likelihood of minimizing car use
5. Likelihood of purchasing green products
6  Likelihood of reducing hydro usage
7. Likelihood of reducing waste by repurposing used products
8. Likelihood of sorting waste correctly
9. Likelihood of walking or cycling shorter distances

Each model is built using the same structure but predicts a different outcome variable. This approach allows us to analyze the specific impact of age and education on each climate-friendly action.

### Likelihood of investing in an electric or hybrid vehicle
@model-one
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-one
#| fig-cap: Probability tree for likelihood of investing in an electric or hybrid vehicle, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/vehicle_electric_tree.png"))

### Likelihood of making home improvements
@model-two
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-two
#| fig-cap: Probability tree for likelihood of making home improvements predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/home_improvement_tree.png"))

### Likelihood of adopting meat alternatives
@model-three
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-three
#| fig-cap: Probability tree for likelihood of adopting meat alternatives, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/protein_alternative_tree.png"))

### Likelihood of minimizing car use
@model-four
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-four
#| fig-cap: Probability tree for likelihood of minimizing car use, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/minimize_car_tree.png"))

### Likelihood of purchasing green products
@model-five
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-five
#| fig-cap: Probability tree for likelihood of purchasing green products, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/green_product_tree.png"))

### Likelihood of reducing hydro usage
@model-six
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-six
#| fig-cap: Probability tree for likelihood of reducing hydro usage, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/reduce_hydro_tree.png"))

###  Likelihood of reducing waste
@model-seven
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-seven
#| fig-cap: Probability tree for likelihood of reducing waste, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/reduce_waste_tree.png"))

###  Likelihood of sorting waste correctly
@model-eight
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-eight
#| fig-cap: Probability tree for likelihood of sorting waste correctly, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/sort_waste_tree.png"))

###  Likelihood of walking or cycling shorter distances
@model-nine
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: model-nine
#| fig-cap: Probability tree for likelihood of walking or cycling shorter distances, predicted by age and education.

knitr::include_graphics(here::here("data/04-model_data/short_distance_tree.png"))

## Model Estimation and Interpretation ##
The probability tree model estimates how the predictors age and education influence the likelihood of adopting climate-friendly behaviors. By splitting the data into branches based on these variables, the model helps identify decision rules that predict individuals' likelihood of taking actions such as reducing waste or investing in green products. We anticipate that younger individuals, due to greater environmental awareness, are more likely to engage in climate-positive actions, while those with higher levels of education may also be more inclined to adopt such behaviors, given their increased awareness of climate change and sustainable practices. The model estimates the parameters for age and education using regression techniques, providing coefficients that indicate how each factor influences the likelihood of taking climate-friendly actions.
